---
title: "Modelling Cognitive Decline"
author: "Bindoff, A."
date: "created: 4 July 2018"
output: ioslides_presentation
autosize: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(reshape2)
library(dplyr)
library(ggplot2)
library(lme4)
library(purrr)
library(brms)
library(gridExtra)
library(viridis)
```

## Mild Cognitive Impairment (MCI)

- classification typically used in prospective studies of older adults
- identify participants who are more likely to receive proximal diagnosis of dementia
- increasingly used as a **clinical diagnosis**

## Diagnostic criteria

- sex- and age-adjusted z-scores from cognitive test/s
- typically rely on a threshold; e.g one or more tests with $z < 1.5$
- false positive rate vulnerable to inflation
  - within-subject variation
  - multiple testing
  - population heterogeneity
  
## Change of states

- prodromal phase of dementia indicating developing pathological state?
- classification/diagnosis only accurate if *acyclic*
  - no recovery, only progression to dementia

## Improving classification/diagnosis

- reducing error will improve power of studies
- proposed solutions to date -
  - "raise the bar"
    - more stringent thresholds
    - differentiating between scores on cognitive domains (which may not be orthogonal)
  - state-space models?
  
## Proposed solution

- wish list:
  - single-subject or multiple-subject designs
  - single-test or multiple-test designs (*that control Type 1 error!*)
  - *must model a change from cognitively healthy to cognitive decline state*

## Piece-wise linear regression?

- assume cognitively normal at beginning of study
- allow possibility of cognitive decline at some point
- fit two slopes to data
- find **break-point** that best fits
  1. **data**
  2. **a process of transition from healthy to cognitive decline!**
  


---
```{r, echo = F}
set.seed(1)

tc <- function(x,t){
  c(rep(0, x[3]), ((x[3]+1)-x[3]):(t-x[3]))*x[4]+ rnorm(t, x[1], sqrt(x[2]))
}

sim <- function(id, j = 9, t = 12, j.mu = 0, j.var = 1){
  s <- data.frame(j.mu = j.mu,
                  j.var = j.var,
                  cp.t = sample(3:t-2, 1),
                  cp.slope = -abs(rnorm(j, 0, 0.66)))
  p <- melt(data.frame(apply(s, 1, function(x) tc(x, t))))
  p$tau <- as.double(rep(c(1:t), j))
  p$id <- id
  p$cp <- s$cp.t
  names(p) <- c("test", "score", "tau", "id", "cpT")
  p
}  

j = 9
j.mu = rnorm(j, 0, 1)
j.var = rnorm(j, 0, 0.5)^2
  
p <- sim("A", j.mu = j.mu, j.var = j.var)
p$cp <- 0


```

```{r, echo = FALSE}
i_fit <- function(p, i){
  p$cp <- ifelse(p$tau > i, p$tau - i, 0)
  p$fit <- predict(lm(score ~ tau + cp, data = p), p)
  p$facet <- i
  p
}

p0 <- lapply(1:12, function(x) i_fit(p, x))
p0 <- bind_rows(p0)


ggplot(p0, aes(x = tau, y = score, colour = test, xintercept = facet)) +
  geom_line(size = 3/4) +
  scale_colour_viridis(discrete = TRUE) +
  facet_wrap(~facet) +
  geom_line(aes(y = fit), colour = "red", size = 1) +
  geom_vline(data = p0, aes(xintercept = facet), linetype = "dashed") +
  xlab("Testing phase") +
  scale_x_continuous(breaks = c(1,3,5,7,9,11)) +
  theme(panel.background = element_blank(),
        axis.line = element_line()) 


```

---

```{r, echo = TRUE}
loss <- function(i){
  p$cp0 <- ifelse(p$tau > i, p$tau - i, 0)
  deviance(lmer(score ~ tau + cp0 + (tau + cp0|test), data = p))
}

which.min(sapply(unique(p$tau), loss))
```
## Problems

Fitting the data is easy (OLS, maximum-likelihood), but maximising $p(y|m_i)$ is not guaranteed to find the break-point that marks the beginning of cognitive decline (if cognitive decline is even present).

---

```{r, echo = F, message = F, warning = F}

set.seed(1)


tc0 <- function(x){
  c(x[3]*c(1,2,3), x[4]*c(1,2,3), x[5]*c(1,2,3)) + rnorm(9, x[1], sqrt(x[2]))
}


sim0 <- function(id, j = 9, t = 9, j.mu = 0, j.var = 1){  # change-point can lead to positive or negative slope
  s <- data.frame(j.mu = j.mu,
                  j.var = j.var,
                  slope1 = sample(c(-1, 1), 1)*abs(rnorm(j, 0, 3/4)),
                  slope2 = sample(c(-1, 1), 1)*abs(rnorm(j, 0, 3/4)),
                  slope3 = sample(c(-1, 1), 1)*abs(rnorm(j, 0, 3/4)))
  p <- melt(data.frame(apply(s, 1, function(x) tc0(x))))
  p$tau <- as.double(rep(c(1:t), j))
  p$id <- id
  p$cp <- 0
  names(p) <- c("test", "score", "tau", "id", "cpT")
  p
}  
t = 12
j = 9
j.mu = rnorm(j, 0, 1)
j.var = rnorm(j, 0, .5)^2
#j.var = sample(c(0.01, 2), j, replace = T)

p0 <- map(c("A", "B", "C", "D", "E", "F"), function(x) sim0(x, j.mu = j.mu, j.var = j.var))



loss <- function(i, p){
  p$cp0 <- ifelse(p$tau > i, p$tau - i, 0)
  deviance(lmer(score ~ tau + cp0 + (tau + cp0|test), data = p))
}


cp_min <- unlist(lapply(p0, function(p) which.min(sapply(1:9, function(i) loss(i, p)))))


p0 <- bind_rows(p0)
p0$cp_min[p0$id == "A"] <- as.numeric(cp_min[1])
p0$cp_min[p0$id == "B"] <- as.numeric(cp_min[2])
p0$cp_min[p0$id == "C"] <- as.numeric(cp_min[3])
p0$cp_min[p0$id == "D"] <- as.numeric(cp_min[4])
p0$cp_min[p0$id == "E"] <- as.numeric(cp_min[5])
p0$cp_min[p0$id == "F"] <- as.numeric(cp_min[6])


ggplot(p0, aes(x = tau, y = score, colour = test)) +
  geom_line(size = 3/4) +
  scale_colour_viridis(discrete = TRUE) +
  facet_wrap(~id) +
  #geom_line(aes(y = fit), colour = "red", size = 1) +
  geom_vline(data = p0, aes(xintercept = cp_min), linetype = "dashed") +
  xlab("Testing phase") +
  scale_x_continuous(breaks = c(2,4,6,8)) +
  theme(panel.background = element_blank(),
        axis.line = element_line()) 
```

## Modelling transition

We're not really interested in finding the model that is *best explained by* the data, we're really interested in finding the model that *best explains* the data **if** the data are explained by cognitive decline.  

Ideally this model will have $\beta_1 \approx 0$ when the subject is in a healthy state and $\beta_2 + \beta_1 \leq 0$ when the subject is experiencing cognitive decline.

We seek to maximise $Pr(m_i|y)$ where $Pr(m_i|y) = \frac{Pr(y|m_i)Pr(m_i)}{\sum_{j=1}^{m} Pr(y|m_j)Pr(m_j)}$, $\Pr(y|m_i)$ is the likelihood and $Pr(m_i)$ is a prior probability that represents our beliefs about $\beta_{1i}, \beta_{2i}$.

---

```{r, hold = TRUE}
loss_p <- function(i){
  p$cp0 <- ifelse(p$tau > i, p$tau - i, 0)
  lmer(score ~ tau + cp0 + (tau + cp0|test), data = p)
}

prior_b1 <- function(x, mean = 0, sd = 1/5){
  dnorm(x, mean = mean, sd = sd, log = TRUE)
}

prior_b2 <- function(x, mean = -1, sd = 1/5){
  dnorm(x, mean = mean, sd = sd, log = TRUE)
}

g <- data.frame(tau.lwr = double(), 
                tau.upr = double(),
                cp.lwr = double(), 
                cp.upr = double(),
                cp = integer(),
                id = character(),
                row.names = NULL)

for(k in unique(p0$id)){
  p <- filter(p0, id == k)
  
  l <- sapply(2:11, loss_p)

  cp_min <- which.max(sapply(l, function(x) exp(prior_b1(fixef(x)[2L])+prior_b2(fixef(x)[3L]+fixef(x)[2L])+logLik(x))))+1
  
  p$cp <- ifelse(p$tau > cp_min, p$tau - cp_min, 0)
  m1 <- lmer(score ~ tau + cp + (tau + cp|test), data = p)
  m2 <- lm(score ~ tau + cp, data = p)
  p0$fit[p0$id == k] <- predict(m2, p)
  p0$cp_min[p0$id == k] <- cp_min
  #cat("Subject ", k, "  \nestimated cp = ", cp_min, "  \n")
  ci <- confint(m1, method = "Wald")[8:10,]
  #print(ci)
  g <- rbind(g, data.frame(tau.lwr = ci[2,1], tau.upr = ci[2,2], cp.lwr = ci[3,1], cp.upr = ci[3,2], cp = cp_min, id = k, row.names = NULL))
}

```

```{r}
g$cog_dec <- g$tau.lwr <= 0 & g$tau.upr>=0 & g$cp.lwr <0 & g$cp.upr <= 0
g
```

---

```{r}
ggplot(p0, aes(x = tau, y = score, colour = test)) +
  geom_line(size = 3/4) +
  scale_colour_viridis(discrete = TRUE) +
  facet_wrap(~id) +
  geom_line(aes(y = fit), colour = "red", size = 1) +
  geom_vline(data = p0, aes(xintercept = cp_min), linetype = "dashed") +
  xlab("Testing phase") +
  scale_x_continuous(breaks = c(1,3,5,7,9,11)) +
  theme(panel.background = element_blank(),
        axis.line = element_line()) 

```


```{r, eval = F}
library(merTools)
p <- filter(p0, id == "E")
p$cp0 <- ifelse(p$tau > 6, p$tau - 6, 0)
m1 <- lmer(score ~ tau + cp0 + (tau + cp0|test), data = p)

ran <- REsim(m1, n.sims = 1000)
plotREsim(ran, level = .95)
```